### 2014-04-18 text/Chapter03.tex
Working on structural liveness (Chapter 3)

\subsection{Scalars and vectors}

@@

It is useful to clarify how vectors are used in this thesis. Suppose we have a set $\C$ of
\emph{scalars} such that $\cclrd{r_1},\ldots,\cclrd{r_n} \in \C$. A vector $\aclrd{R}$ 
over $\C$ is a tuple $\alift{ \cclrd{r_1}, \ldots, \cclrd{r_n} }$ of scalars. 
We use upper-case letters like $\aclrd{R}, \aclrd{S}, \aclrd{T}$ for vectors and lower-case
letters $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars\footnote{For better readability, the thesis
also distinguishes different structures using colours. However ignoring the colour does not introduce 
any ambiguity.}. We also say that a \emph{shape} of a vector $\slift{\aclrd{R}}$ (or more generally any container) 
is the set of \emph{positions} in a vector. So, a vector of length $n$ has shape $\{ 1, 2, \ldots, n \}$. 
We discuss containers and shapes further in Chapter~\ref{ch:unified} and also discuss how our use
relates to containers of Abbott, Altenkirch and Ghani \cite{types-containers}.

Just as in the usual multiplication of a vector by scalar, we lift any binary operation on scalars into a 
scalar-vector one. For any binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
 $\cclrd{s} \circ \aclrd{R} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{R}, \aclrd{S}$ of the
same shape with positions $\{ 1, \ldots, n \}$ and a relation $\varpropto\, \subseteq \C \times \C$ we define 
$\aclrd{R} \varpropto \aclrd{S} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
Finally, we often concatenate vectors -- for example, when joining two variable contexts.
Given vectors $\aclrd{R}, \aclrd{S}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and 
$\{ 1, \ldots, m \}$, the associative operation for concatenation $\times$ is defined as 
$\aclrd{R}\times\aclrd{S} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.

@@

We now turn our attention to system where additional contextual information are associated not
with the context as a whole (or program scope), but with individual variables. We start by looking
simple static analysis -- variable \emph{liveness}. Then we revisit data-flow computations and
look at applications in security and software updating.

% % % % % % % %

Coeffects are way to describe notions of context in programming that keep turning up. 
To illustrate this, we overview three systems tracking contextual properties that 
 motivate our general coeffect system. Two systems track per-variable properties (bounded 
linear logic and dataflow) and one tracks whole-context properties (implicit parameters).

@@

\subsection{Liveness}
\label{sec:applications-struct-live}

todo

@@

@@

\begin{figure}[t]
\vspace{-0.5em}
\begin{align*}
\begin{array}{c}
%
\hspace{-0.6em}
\begin{array}{c}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{1}} \vdash x : \tau} 
\quad 
\tyruleCompact{weak}
  {\coctx{\Gamma}{R} \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{R} \cons \alift{0}} \vdash e : \tau} 
\\[1.8em]
\tyruleCompact{sub}
  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{R'}} \vdash e : \tau}~(\text{\scriptsize $\aclrd{R} \leq \aclrd{R'}$})
\quad 
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \sigma}{\aclrd{R} \cons \alift{s}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau} 
\\[1.2em]
\end{array} \\
%
% ------------------------ 
%
\begin{array}{c}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau \quad 
    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} \ast \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
\\[1.3em]
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \subst{e}{z,y}{x} : \tau}
\\[1.3em]
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{Q}} \vdash e : \tau}
\end{array} 
% ------------------------ 
%
% \hspace{-1em}
%\begin{array}{c}
%
% TP: I removed the 'let' rule again because we do not need it to explain the storage axiom here.
% Also, I think the BLL sample is already used to explain quite a lot of things, so I'd be happier
% to keep it simple.
%
% BTW: The \tyruleCompact macro creates a rule that *looks* differently..
% (I do not understand why but it makes the line somehow thicker...)
%
% \tyruleCompact{let}
%   {\coctxs{\Gamma_1, v : \sigma}{R \times \langle{t}\rangle} \vdash e_1 : \tau \quad 
%    \coctxs{\Gamma_2}{S} \vdash e_2 : \sigma}
%   {\coctxs{\Gamma_1, \Gamma_2}{R \times (t \ast S)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau} \\[1.8em]
\end{array}
\end{align*}
\vspace{-0.5em}
\caption{type and coeffect system for bounded reuse}
\label{fig:bounded-coeff}
\end{figure}

>>>>

\subsection{Scalars and vectors}
\label{sec:applications-strucutre-vec}

@@

It is useful to clarify how vectors are used in this thesis. Suppose we have a set $\C$ of
\emph{scalars} such that $\cclrd{r_1},\ldots,\cclrd{r_n} \in \C$. A vector $\aclrd{R}$ 
over $\C$ is a tuple $\alift{ \cclrd{r_1}, \ldots, \cclrd{r_n} }$ of scalars. 
We use bold-face letters like $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ for vectors and lower-case
letters $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars\footnote{For better readability, the thesis
also distinguishes different structures using colours. However ignoring the colour does not introduce 
any ambiguity.}. We also say that a \emph{shape} of a vector $\slift{\aclrd{\textbf{r}}}$ (or more generally any container) 
is the set of \emph{positions} in a vector. So, a vector of length $n$ has shape $\{ 1, 2, \ldots, n \}$. 
We discuss containers and shapes further in Chapter~\ref{ch:unified} and also discuss how our use
relates to containers of Abbott, Altenkirch and Ghani \cite{types-containers}.

Just as in the usual multiplication of a vector by scalar, we lift any binary operation on scalars into a 
scalar-vector one. For any binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
 $\cclrd{s} \circ \aclrd{\textbf{r}} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ of the
same shape with positions $\{ 1, \ldots, n \}$ and a relation $\varpropto\, \subseteq \C \times \C$ we define 
$\aclrd{\textbf{r}} \varpropto \aclrd{\textbf{s}} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
Finally, we often concatenate vectors -- for example, when joining two variable contexts.
Given vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and 
$\{ 1, \ldots, m \}$, the associative operation for concatenation $\times$ is defined as 
$\aclrd{\textbf{r}}\times\aclrd{\textbf{s}} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.

@@

In structural coeffect systems, the additional information are associated with individual variables.
This is very often information about how the variables are used, or, in which contexts they are 
used.

In Chapter~\ref{ch:intro}, we introduced the idea using an example that tracks array access patterns.  
Each variable is annotated with a range specifying which elements of the corresponding array 
may be accessed. In this section, we look at a number of examples. We first consider an example
inspired by linear logic. Then we revisit liveness and data-flow, for which the structural system 
provides a more precise analysis. Finally, we look at a number of other practical uses including 
security, tainting and provenance tracking.

@@

\subsection{Liveness analysis revisited}
\label{sec:applications-struct-live}

The flat system for liveness analysis presented in Section~\ref{sec:applications-flat-live} is 
interesting from a theoretical perspective, but it is not practically useful. In this section, we
revisit the problem and define a structural system that tracks liveness per-variable.

\paragraph{Structural liveness.} 
Recall two examples discussed earlier where the flat liveness analysis marked the whole context
as (syntactically) live, despite the fact part of it was (semantically) dead:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{constant} = \lambda y \rightarrow \lambda x \rightarrow y\\
\kvd{let}~\ident{answer} = (\lambda x \rightarrow x)~42
\end{array}
\end{equation*}
%
In the first case, the variable $x$ is dead, but was marked as live. In the second example, the
declaration-site of the \ident{answer} value is dead, but was marked as live. This is because in 
both of the expressions, \emph{some} variable is accessed. However, the (\emph{abs}) rule of flat
liveness has no way of determining \emph{which} variables are used by the body -- and, in particular,
whether the accessed variable is the \emph{bound} variable or some of the \emph{free} variables.

As discussed earlier, we can resolve this by attaching a \emph{vector} of liveness annotations to
a \emph{vector} of variables. In the first example, the available variables are $y$ and $x$, so
the variable context $\Gamma$ is a vector $\langle y\!:\!\tau, x\!:\!\tau \rangle$. Only the variable $y$
is used and so the annotated context is: $\coctx{y\!:\!\tau, x\!:\!\tau}{ \alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}} }$.
When writing the contexts, we omit angle brackets around variables, but it should still be viewed
as a vector. There are two important points:

\begin{itemize}
\item The fact that variables are now a vector means that we cannot freely re-order them. This
  guarantees that $\coctx{x\!:\!\tau, y\!:\!\tau}{\alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}}}$
  can not be confused with $\coctx{y\!:\!\tau, x\!:\!\tau}{\alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}}}$.
  We need to define the type system in a way that is similar to sub-structural systems 
  (discussed in Section~\ref{sec:path-logic}) which provide explicit rules for manipulating
  the context.

\item We choose to attach a vector of annotations to a vector of variables, rather than attaching
  individual annotations to individual variables. This lets us unify and combine flat and 
  structural systems as discussed in Chapter~\ref{ch:unified}, but the alternative is briefly
  explored in Chapter~\ref{ch:meta}.
\end{itemize}

\paragraph{Type system.}
The structural system for liveness uses the same two-point lattice of annotations
$\mathcal{L}=\{ \ident{L}, \ident{D} \}$ that was used by the flat system. We also use the
$\sqcup, \sqcap$ and $\sqsubseteq$ operators that are defined in Figure~\ref{fig:applications-flat-livealg}.

The rules of the system are split into two groups. Figure~\ref{fig:applications-struct-live} (a) shows 
the standard syntax-driven rules plus sub-coeffecting. In (\emph{var}), the context contains just the 
single accessed variable, which is annotated as live. Other variables can be introduced using weakening. 
A constant (\emph{const}) is accessed in an empty context, which also carries no annotations. The
sub-coeffecting rule (\emph{sub}) uses a point-wise extension of the $\sqsubseteq$ relation over two
vectors as defined in Section~\ref{sec:applications-strucutre-vec}.

In the (\emph{abs}) rule, the variable context of the body $\Gamma, x\!:\!\tau_1$ is annotated with
a vector $\mathbf{\aclrd{r}}\cons\alift{\cclrd{s}}$, where the vector $\mathbf{\aclrd{r}}$ corresponds
to $\Gamma$ and the singleton annotation $\cclrd{s}$ corresponds to the variable $x$. Thus, the
function is annotated with $\cclrd{s}$. Note that the free-variable context is annotated with vectors,
but functions take only a single input and so are annotated with primitive annotations.

The (\emph{app}) rule is similar to function applications in flat systems, but there is an important
difference. In structural systems, the two sub-expressions have separate variable contexts
$\Gamma_1$ and $\Gamma_2$. Therefore, the composed expression just concatenates the variables
and their corresponding annotations. (We can still use the same variable in both sub-expressions
thanks to the structural contraction rule.)

The context $\Gamma_1$ is used to evaluate $e_1$ and is thus annotated with $\aclrd{\mathbf{r}}$. 
The annotation for $\Gamma_2$ is more interesting. It is a result of sequential composition of two
semantic functions -- the first one takes the (multi-variable) context $\Gamma_2$ and evaluates 
$e_2$; the second takes the result of type $\tau_1$ and passes it to the function $\tau_1 \xrightarrow{\cclrd{t}} \tau_2$.
The composition is defined as follows:
%
\begin{equation*}
g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\aclrd{\mathbf{s}}} \sigma
\qquad
f : \sigma \xrightarrow{\cclrd{t}} \tau
\qquad
f \circ g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\cclrd{t} \sqcup \aclrd{\mathbf{s}}} \tau
\end{equation*}
%
This definition is only for illustration and is revised in Chapter~\ref{ch:structural}. The function
$g$ takes a product of multiple variables (and is annotated with a vector). The function $f$ takes
just a single value and is annotated with the scalar. As in the flat system, sequential composition
is modelled using $\sqcup$ -- but here, we use a scalar-vector extension of the operation. Finally,
the (\emph{let}) rule follows similar reasoning (and also corresponds to the typing of $(\lambda x.e_2)~e_1$.

\paragraph{Structural typing rules.}

The rules are split into two groups. We first discuss the standard rules.

is good

@@

\begin{figure}[t]
{\small a.) Ordinary, syntax-driven rules with sub-coeffecting}
\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{const}
  {c:\tau \in \Delta}
  {\coctx{()}{\alift{}} \vdash c : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \sqsubseteq \aclrd{\textbf{r'}}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, x : \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
\end{equation*}
\vspace{0.5em}

{\small b.) Structural rules for context manipulation}
\begin{equation*}
\tyrule{weak}
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{0}} \vdash e : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
\end{equation*}
\begin{equation*}
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
\end{equation*}

\caption{Structural coeffect liveness analysis}
\label{fig:applications-struct-live}
\vspace{-1em}
\end{figure}

@@

\begin{figure}[t]
{\small Ordinary rules}

\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{1}} \vdash x : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau} 
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \sigma}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, v : \sigma}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \sigma}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau}
\end{equation*}

{\small Structural rules}

\begin{equation*}
\tyrule{weak}
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{0}} \vdash e : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
\end{equation*}
\begin{equation*}
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
\end{equation*}

\caption{Coeffect rules for tracking implicit parameters}
\label{fig:applications-struct-bll}
\vspace{-1em}
\end{figure}
